---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Cost of hosting world cup

Let's start by loading the libraries that we'll utilize in our analysis

```{r message=FALSE, warning=FALSE}
#web scrapping
library(rvest)
library(robotstxt)
#
library(tidyverse)
#handling spatial-data 
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(ggflags)
library(ggspatial)
library(giscoR)
library(rasterpic)
library(countrycode)
library("dataverse")
#adding flags in ggplot
library(ggimage)
#Visually explore data tables 
library(visdat)
#fit text within a defined area
library(ggfittext)
#set the default ggplot theme
theme_set(cowplot::theme_cowplot())
#get image
library(ggwordcloud)
```

Define variables...
```{r}
#source of scraped data
url_2 <- "https://en.wikipedia.org/wiki/Economics_of_the_FIFA_World_Cup"
robotstxt::paths_allowed(url_2)#check if the webpage is accessible for scraping
# get flags form this repo
flagrepo <- "https://raw.githubusercontent.com/hjnilsson/country-flags/master/png250px/"
#world cup image
wc_img <- here::here("images/wc12.png")

#
url3 <- "https://en.wikipedia.org/wiki/Cost_of_the_Olympic_Games"
```

and functions.
```{r}
#define function to control transparency
transparent_0.1 <- function(img) {
  magick::image_fx(img, expression = glue::glue("0.1*a"), channel = "alpha")
}
transparent_0.2 <- function(img) {
  magick::image_fx(img, expression = glue::glue("0.2*a"), channel = "alpha")
}
```

Scrape data from wikipedia article
```{r}
# get the Costs of World Cups
df_4 <- url_2 %>%
  xml2::read_html() %>% 
  html_element(".wikitable") %>% # select table with data of interest
  html_table()

```

Clean and prepare data for the anlaysis
```{r}
#
df_4 <- df_4 %>%
  janitor::clean_names() %>% #clean column names
  mutate(host = str_replace(host,  "/\\s", "-")) %>% 
  filter(general_cost != "n.a.") %>% #remove world cup with unknown cost
  extract(host, c("country","year"), "([A-Z-]+) (\\([0-9]+\\))") %>% #put country name and year into separate columns
  separate_rows(country, sep = "-") %>% #separate cohosts into individual rows
  mutate(year = parse_number(year),
         x =  parse_number(general_cost),#get cost in numbers
         cost = ifelse(str_detect(general_cost, "million"),x/1000,x),#cost in billions
         x = NULL,
         general_cost = NULL)

#add country iso2c code
df_4$iso2 <- countrycode(df_4$country, "ioc", "iso2c")
```

Bar plot of cost of hosting world cup
```{r}
ggplot()+
  geom_col(data = df_4 %>% distinct(year, cost),
           aes(year, cost),
           fill = "grey90")+
  geom_text(data = df_4 %>% distinct(year, cost),
            aes(year, cost, label = cost),
            size = 5, nudge_y = 8)+
  scale_y_continuous(breaks = c(1,10,100,200))+
  labs(y = "Billion $")
```

Add country flag on top of bars, fill bars with dollar signs, and add image to background.

```{r}
ggplot()+
  #basic bar for each worldcup
  geom_col(data = df_4 %>% distinct(year, cost),
           aes(year, cost))+
  #add cost on top of each bar
  geom_text(data = df_4 %>% 
              distinct(year, cost) %>%
              mutate(y2 =cost, y = glue::glue("${cost} bn")),
            aes(year, y2, label = y),
            size = 4, nudge_y = 20)+
  #add country flags on top of the bars but below the cost
  ggimage::geom_flag(data = df_4 %>%
                       filter(!country %in% c("JPN", "KOR")),
                     aes(year, cost+8, image = iso2),
                     size = 0.06)+
  #the same as the previous step but for flags of cohosting countries
  ggimage::geom_flag(data = df_4 %>%
                       filter(country %in% c("JPN", "KOR")) %>%
                       mutate(year=ifelse(str_detect(country, "JPN"), year+1.5, year)),
                     aes(year-0.7, cost+8, image = iso2),
                     size = 0.035)+
  #fill bars with dollar signs
  ggpattern::geom_col_pattern(data = df_4 %>% distinct(year, cost),
                               aes(year, cost),
                               pattern_filename = I("images/Gold-Dollar-Sign-PNG-HD.png"), 
                               pattern = 'image', 
                               pattern_type = 'tile',
                               pattern_scale = -1,
                               fill = "#a1d99b")+
  #add world cup image to the background
    ggimage::geom_image(data = data.frame(x = 1995, y = 120),
                      aes(x,y),
                      image = wc_img,image_fun = transparent_0.2,
                      size = 1.2)+
  #control x and y axis
      scale_x_continuous(breaks = full_seq(range(df_4$year), 4))+
  scale_y_continuous(breaks = c(1,10,100,200))+
  labs(y = "", x = "")+
  #control plot theme
    theme(plot.background = element_rect(fill =  "#e5f5e0"),
          panel.background = element_rect(fill =  "#e5f5e0"),
          panel.grid.major = element_line(colour = "#e5f5e0"),
          axis.line = element_blank(),
          axis.ticks = element_blank(),
          axis.text.y = element_blank())
```


```{r}
flags <- c()
# Loop and add
for (iso in df_4$iso2) {
  # Download pic and plot
  imgurl <- paste0(flagrepo, tolower(iso), ".png")
  tmpfile <- tempfile(fileext = ".png")
  download.file(imgurl, tmpfile, quiet = TRUE, mode = "wb")
  
  flags <- c(flags, tmpfile)
}

```

We can do the opposite and instead fill the bars with the country flag
```{r}
ggplot()+
    ggimage::geom_image(data = data.frame(x = 1995, y = 30),
                    aes(x,y),
                    image = wc_img,
                    image_fun = transparent_0.1,
                    size = 1.2)+
    ggimage::geom_image(data = df_4 %>%
                          distinct(year, cost) %>%
                          mutate(y2 = cost+(cost*0.85)) ,
                        aes(year, y2),
                        size = df_4 %>% distinct(year, cost) %>% mutate(x = ceiling(log10(cost*10))/10) %>% pull(x),
                    image = "images/Gold-Dollar-Sign-PNG-HD.png",
                    image_fun = transparent_0.2)+
   ggpattern::geom_col_pattern(data = df_4 ,
                               aes(year, cost, fill = country),
                               position = position_dodge(),
                               pattern_gravity = I("West"),
                               pattern_filename = I(flags), 
                               pattern = 'image', 
                               pattern_type = 'tile',
                               pattern_scale = -1,
                               show.legend = FALSE)+
  geom_text(data = df_4 %>%
              distinct(year, cost) %>%
              mutate(y2 = cost+(cost*0.45),
                     y = glue::glue("$cost} bn")),
            aes(year, y2, label = cost),
            size = 4)+
  geom_text(data = df_4%>%
              arrange(country) %>%
              group_by(year) %>%
              summarise(iso2 = paste(iso2, collapse = "-")),
          aes(year, label = iso2),
          y = -0.07,
          size = 4)+
  scale_y_continuous(breaks = c(1,10,100,200),
                   trans = scales::pseudo_log_trans(base = 10),
                   expand = expansion(add = c(0.2,0.5))
                   )+
  scale_x_continuous(breaks = unique(df_4$year))+
  labs(y = "", x = "")+
  theme(plot.background = element_rect(fill =  "#e5f5e0"),
      panel.background = element_rect(fill =  "#e5f5e0"),
      panel.grid.major = element_line(colour = "#e5f5e0"),
      axis.line = element_blank(),
      axis.ticks = element_blank(),
      axis.text.y = element_blank(),
      axis.text.x = element_text(size = 15))
```

country flag as points connected by a line and put y axis on log10 scale
```{r}
ggplot()+
  geom_line(data = df_4 %>% distinct(year, cost),
            aes(year, cost),
            color = "white",
            size = 1)+
  ggflags::geom_flag(data = df_4,
                     aes(year, cost, country = tolower(iso2)),
                     size = 9)+
  scale_y_continuous(breaks = c(1,10,100,200),
                     trans = scales::pseudo_log_trans(base = 10),
                     expand = expansion(add = c(0.5,0.5))
                     )+
    scale_x_continuous(breaks = unique(df_4$year))+
  geom_text(data = df_4 %>% distinct(year, cost) %>% mutate(y2 = cost+(cost*0.45)+1, y = glue::glue("${cost} bn")),
            aes(year, y2, label = y),
            size = 4)+
    geom_text(data = df_4%>%
                arrange(country) %>%
                group_by(year) %>%
                summarise(iso2 = paste(iso2, collapse = "-"),
                          cost = unique(cost)) %>% 
                ungroup() %>% 
                mutate(y2 = cost-(cost*0.45)-1),
            aes(year,y2 ,  label = iso2),
            size = 4)+
  labs(y = "", x = "")+
    theme(plot.background = element_rect(fill =  "#e5f5e0"),
        panel.background = element_rect(fill =  "#e5f5e0"),
        panel.grid.major = element_line(colour = "#e5f5e0"),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.text.y = element_blank())
```


```{r eval=FALSE, include=FALSE}
#fill world cup with dollar signs

#https://lepennec.github.io/ggwordcloud/articles/ggwordcloud.html
img_path <- "images/png-transparent-fifa-world-cup-trophy-vince-lombardi-trophy-award-soccer-trophy-sport-medal-black.png"
ggplot() +
  geom_text_wordcloud_area(data = love_words,
                           aes(size = speakers),
                           color = "#31a354",
                           label = "$",
                           mask = png::readPNG(img_path),
                           eccentricity = 2,
                           rstep = 0.001,
                           tstep = 0.01,
                           perc_step = 0.01,
                           grid_size = 0.1,
                           grid_margin = 2,
                           rm_outside = TRUE
                           ) +
  scale_size_area(max_size = 40) +
  theme_minimal() 
ggsave("images/worldcloud_dollar.png")
```


Data on cost of sports events (Olympics, worldcup)

Retreive data
```{r}
#https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/CPQEHN
wc_server <- "dataverse.harvard.edu"
wc_doi <- "10.7910/DVN/CPQEHN"
wc_file <- dataverse::dataset_files(dataset = wc_doi, 
                                    server = wc_server)

wc_data_raw <- energy <- dataverse::get_dataframe_by_name(
  filename = wc_file[[1]]$label,
  .f = readxl::read_xlsx,
  original = TRUE,
  dataset = wc_doi, 
  server = wc_server)
```

clean data and make it ready for analysis
```{r}
wc_data <- wc_data_raw %>%
  janitor::clean_names()
wc_data <- wc_data[-1,-2]
wc_data <- wc_data %>%
  tidyr::fill(x1, .direction = "down") %>% #fill missing values with previous ones
  dplyr::rename(event_type = "x1") %>% 
  dplyr::select(!starts_with("source")) %>% #remove columns starts with "source"
  tidyr::separate(x3, into = c("year", "host"), sep = 5) %>% #separate x3 at the fifth position into two new columns
  tidyr::extract(host, into = "host" , regex = "([[:alpha:]\\s\\&]+)") %>% #extract host name (includes letters, spaces, and &)
  mutate(host = str_trim(host), #remove trailing whitespaces
         year = as.integer(year)) %>% 
  mutate(across(matches("cost|number|revenue"), as.numeric)) #convert columns into numeric
```

```{r}
wc_data <- wc_data %>%
  mutate(host = case_when(
    host == "West Germany" ~ "Germany",
    host == "England" ~ "UK",
    TRUE ~ as.character(host)), #rename country name when needed
    host_type = ifelse(host %in% maps::world.cities$country.etc, "country", "city"),#is the host name a city or a country?
    host_type = ifelse(host == "Japan & South Korea", "country", host_type)) %>%
  left_join(maps::world.cities , by = c("host"="name")) %>% #merge the data with table that connects country with city
  mutate(host_country = ifelse(host_type=="country", host, country.etc), #set host country name
         host_city = ifelse(host_type=="city", host, NA), #set host city name
         host_country = case_when(
           host_city == "PyeongChang" ~ "South Korea",
           host_city == "Japan & South Korea" ~ "Japan & South Korea",
           host_city =="Lake Placid" ~ "USA",
           TRUE ~ as.character(host_country)
           )) %>%
  pivot_longer(matches("cost|revenue"),
               names_to = "cash_source",
               values_to = "cash_value") %>%
  mutate(cash_type = str_extract(cash_source, "cost|revenue"),
         cash_value = cash_value/1e9)
```


```{r}
# get the Costs of World Cups
df_5 <- url3 %>%
  xml2::read_html() %>% 
  html_element(".wikitable") %>% # select table with data of interest
  html_table()

```

